Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-11-02 demystify.tex
updates

\chapter*{Preface}
Having been involved in PhD admission committees for many years, I've realized that many \textbf{international} students, especially those in smaller countries or less well-known universities, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that getting admitted to a CS PhD program in the US is difficult compared to other countries.

@@

\paragraph{Not a Good fit and Bad luck}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you will not be admitted.
Similarly, if no faculty is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

@@

\paragraph{You aim too high} This is a very common reason: you have applied to schools that are too high for your profile. While being ambitious is good, you should understand how PhD admission works (e.g., by reading this document and realizing things such as having a good GPA doesn't mean much to top programs) and be realistic!

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  Profs. might not care much about these, but the college or the university often sets a minimum that you need to pass to be considered, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while profs. are willing to argue for your case, they might be reluctant to go against the requirement of the university.

@@

\paragraph{Red flags} Various types of red flags can cause concern to the adcom (e.g., too many STEM courses with low or withdrawn grades, plagiarism, cheating, or other academic dishonesty). Another one is that you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). Adcom members might have contacts in other programs and heard about your stories, or your LoRs might mention these.

@@

\begin{itemize}
    \item \textbf{Send emails to the wrong prof.}. Most students often do not pay attention or know about this, but a very common reason why you don't get a reply is that you send email to profs. who do not / cannot advise CS PhD students (e.g., adjunct, emeritus, teaching, non-CS, etc.). See \autoref{sec:faculty-types} for details. While these profs. might be able to co-advise, they typically do not have the bandwidth, funding, and the desire to mentor students (they are already overloaded with teaching and other service duties). So do some research before sending emails, e.g., most CS profs. who are active in research will have a website with their work (publications) and students; some also have lab website dedicated to their group. Most teaching and adjunct profs. do not have such a research page.

>>>>

\chapter*{Preface}
Having been involved in PhD admission committees for many years, I've realized that \textbf{international} students, especially those in smaller countries or less well-known universities, often lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that getting admitted to a CS PhD program in the US is difficult compared to other countries.

@@

\paragraph{You aim too high} 
You have applied to schools that are \emph{way too high} for your profile (\autoref{sec:selecting-ranking-schools}). Many students simply just go for very top schools (e.g., top 10) and are surprised when they are rejected to all of them, in multiple years, and completely get shut out.  This is very obvious but many students still do this. An analogy is a person who has never hike wanting to climb Everest, which btw if you could, you might have a better admission chance (\autoref{sec:improve-your-chance}).

While being ambitious and aiming high is good, you should understand how PhD admission works (e.g., by reading \emph{this} document and realizing things such as having a good GPA or GRE doesn't mean much to top programs), that in the US there are many good schools, and just be realistic.

\paragraph{Not a Good fit and Bad luck}  You could have an excellent profile (e.g., strong research and LoRs), but if you are interested in a research area that the program does not have, you will not be admitted.
Similarly, if no faculty is willing to advise you (e.g., they are on sabbatical or personal leave, do not have sufficient funding, or already have too many students), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

@@

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  Profs. might not care much about these, but the college or the university often sets a minimum that you need to pass to be considered, especially for TA funding (a low English score causes concern that you might not be able to communicate well with students as a TA).  Thus, while profs. are willing to argue for your case, they might be reluctant to go against the requirement of the university.

@@

\paragraph{Red flags} Various types of red flags can cause concern to the adcom. Common examples include too many STEM courses with low or withdrawn grades, plagiarism, cheating, or other academic dishonesty. Another one is that you have a history of jumping from one program to another. Adcom members might have contacts in other programs and heard about your stories, or your LoRs might mention these.

@@

\begin{itemize}
    \item \textbf{Send emails to the wrong prof.}. Most students often do not pay attention or know about this, but a very common reason why you don't get a reply is that you send email to profs. who do not / cannot advise CS PhD students (e.g., adjunct, emeritus, teaching, non-CS, etc.). See \autoref{sec:faculty-types} for details. While these profs. might be able to co-advise, they typically do not have the bandwidth, funding, and the desire to mentor students for research (they are already overloaded with teaching and other service duties). So do some homework before sending emails, e.g., most CS profs. who are active in research will have a website with their work (publications) and students; some also have lab website dedicated to their group. Most teaching and adjunct profs. do not have such a research page.
